---
title: Water Consents Dashboard
toc: true
---

## Headline finding

Northland consent records in this dataset are dominated by irrigation and drinking-water uses, with a clear split between surface water and groundwater sources.

## Why this matters for councils

Consented take patterns help identify where pressure on freshwater allocation may increase and where review of expiry profiles, monitoring, or demand-management settings may be most useful.

## Interactive overview

```{r}
#| warning: false
#| message: false
#| code-fold: true
library(readr)
library(dplyr)
library(plotly)
library(sf)
library(leaflet)

consents <- read_csv("../data/consented-freshwater-takes/take_frame.csv", show_col_types = FALSE)

by_use <- consents |>
  count(primaryuse3, sort = TRUE)

plot_ly(
  by_use,
  x = ~reorder(primaryuse3, n),
  y = ~n,
  type = "bar",
  marker = list(color = "#2b7a78")
) |>
  layout(
    xaxis = list(title = "Primary use"),
    yaxis = list(title = "Number of consents"),
    margin = list(b = 120)
  )
```

```{r}
#| warning: false
#| message: false
#| code-fold: true
map_df <- consents |>
  filter(!is.na(easting_nztm), !is.na(northing_nztm)) |>
  mutate(maxannual = suppressWarnings(as.numeric(maxannual)))

pts <- st_as_sf(
  map_df,
  coords = c("northing_nztm", "easting_nztm"),
  crs = 2193,
  remove = FALSE
) |>
  st_transform(4326)

leaflet(pts) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addCircleMarkers(
    radius = 4,
    stroke = FALSE,
    fillOpacity = 0.65,
    color = "#2b7a78",
    popup = ~paste0(
      "Consent: ", resourceconsentid,
      "<br>Use: ", primaryuse3,
      "<br>Source: ", primarysource,
      "<br>Max annual: ", maxannual
    )
  )
```

## Methods and source

- Source: Ministry for the Environment / council-supplied consented freshwater takes.
- Files used: `data/consented-freshwater-takes/take_frame.csv`.
- Spatial reference in source fields: NZTM coordinates.
- Caveat: this page provides an exploratory operational view and does not represent full allocation accounting.

::: {.callout-note collapse="true"}
## Show code

The complete source code is available in folded code chunks above.
:::
