---
title: Freshwater Habitat Quality
toc: true
---

## Headline finding

Most monitored sites in this dataset are rated Fair to Good, with a smaller set of Excellent sites often associated with native or less-modified catchments.

## Why this matters for councils

Habitat-quality indicators support prioritization of restoration investment, catchment action plans, and communication of where ecological gains are being retained or lost.

## Interactive overview

```{r}
#| warning: false
#| message: false
#| code-fold: true
library(readr)
library(dplyr)
library(plotly)
library(leaflet)

habitat <- read_csv("../data/freshwater-physical-habitat/habitat_quality_scores.csv", show_col_types = FALSE)

by_class <- habitat |>
  count(habitat_quality, sort = TRUE)

plot_ly(
  by_class,
  x = ~reorder(habitat_quality, n),
  y = ~n,
  type = "bar",
  marker = list(color = "#ff7a59")
) |>
  layout(
    xaxis = list(title = "Habitat class"),
    yaxis = list(title = "Number of sites")
  )
```

```{r}
#| warning: false
#| message: false
#| code-fold: true
plot_ly(
  habitat,
  x = ~dominant_landcover,
  y = ~habitat_quality_score,
  color = ~habitat_quality,
  colors = c("#ef4444", "#f59e0b", "#10b981"),
  type = "box"
) |>
  layout(
    xaxis = list(title = "Dominant landcover"),
    yaxis = list(title = "Habitat quality score")
  )
```

```{r}
#| warning: false
#| message: false
#| code-fold: true
leaflet(habitat) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addCircleMarkers(
    lng = ~long,
    lat = ~lat,
    radius = 4,
    stroke = FALSE,
    fillOpacity = 0.7,
    color = "#2b7a78",
    popup = ~paste0(
      "Site: ", site_name,
      "<br>Region: ", region,
      "<br>Score: ", habitat_quality_score,
      "<br>Class: ", habitat_quality,
      "<br>Landcover: ", dominant_landcover
    )
  )
```

## Methods and source

- Source: freshwater physical habitat indicator dataset.
- File used: `data/freshwater-physical-habitat/habitat_quality_scores.csv`.
- Indicator fields include score, class, landcover context, and site coordinates.
- Caveat: this is a compact communication layer; interpretation should include local context and sampling design.

::: {.callout-note collapse="true"}
## Show code

The complete source code is available in folded code chunks above.
:::
