---
title: Groundwater Quality State
toc: true
---

## Headline finding

Groundwater quality reporting tracks multiple contaminant indicators over long time horizons, supporting risk screening for drinking-water safety and ecosystem effects.

## Why this matters for councils

Long-term groundwater indicators help identify where sustained pressure may require tighter management, targeted mitigation, or intensified monitoring.

## Interactive overview

```{r}
#| warning: false
#| message: false
#| code-fold: true
library(sf)
library(dplyr)
library(tidyr)
library(plotly)
library(leaflet)

gpkg_path <- "../data/mfe-groundwater/groundwater-quality-state-20102014.gpkg"

layers <- st_layers(gpkg_path)
gw <- st_read(gpkg_path, layer = layers$name[1], quiet = TRUE)

# convert Results columns to numeric type
gw_nodg <- st_drop_geometry(gw)

# create numeric result column from likely field names
candidate_result_names <- c("Result", "result", "Value", "median", "c_median", "ann_max", "q_50", "q_25")
present_result_names <- intersect(candidate_result_names, names(gw_nodg))
if (length(present_result_names) == 0) {
  stop("No candidate result columns found in data; available names: ", paste(names(gw_nodg), collapse = ", "))
}

# build Result_char by coalescing over present columns
gw_nodg$Result_char <- NA_character_
for (nm in present_result_names) {
  gw_nodg$Result_char <- dplyr::coalesce(gw_nodg$Result_char, as.character(gw_nodg[[nm]]))
}
gw_nodg$Result_num <- suppressWarnings(as.numeric(gsub(",", "", gw_nodg$Result_char)))

# detect indicator column name (try a few common variants)
possible_indicator_names <- c(
  "Aggregated_Back_calculated_parameter",
  "aggregated_back_calculated_parameter",
  "Aggregated_Parameter",
  "parameter",
  "Indicator",
  "aggregated_parameter",
  # common abbreviated indicator columns in this dataset
  "np_id",
  "np_id_name",
  "np_id_units"
)
# prefer descriptive `np_id_name` if present, else first match
if ("np_id_name" %in% names(gw_nodg)) {
  indicator_col <- "np_id_name"
} else {
  indicator_col <- intersect(possible_indicator_names, names(gw_nodg))[1]
}
if (is.na(indicator_col) || is.null(indicator_col)) {
  stop("Indicator column not found in data; available names: ", paste(names(gw_nodg), collapse = ", "))
}

quality_profile <- gw_nodg |>
  group_by(.data[[indicator_col]]) |>
  summarise(records = sum(!is.na(Result_num)), .groups = "drop") |>
  rename(indicator = .data[[indicator_col]]) |>
  arrange(desc(records)) |>
  slice_head(n = 12)

plot_ly(
  quality_profile,
  x = ~reorder(indicator, records),
  y = ~records,
  type = "bar",
  marker = list(color = "#2b7a78")
) |>
  layout(
    xaxis = list(title = "Indicator field"),
    yaxis = list(title = "Non-missing records"),
    margin = list(b = 180)
  )
```

```{r}
#| warning: false
#| message: false
#| code-fold: true

# transform dataframe into sf
## Coerce coordinate columns (support lowercase names from updated dataset)
gw <- gw %>%
  mutate(
    Easting = coalesce(as.numeric(easting), as.numeric(easting)),
    Northing = coalesce(as.numeric(northing), as.numeric(northing)),
    long = coalesce(as.numeric(long), as.numeric(long), as.numeric(long)),
    lat = coalesce(as.numeric(lat), as.numeric(lat), as.numeric(lat))
  )

## Build sf either from lon/lat
gw_wgs84 <- gw %>%
  filter(!is.na(long) & !is.na(lat)) %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)


leaflet(gw_wgs84) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addCircleMarkers(
    radius = 4,
    stroke = FALSE,
    fillOpacity = 0.55,
    color = "#ff7a59",
    popup = ~paste0(
      "location: ", s_id,
      "<br>period: ", period,
      "<br>indicator: ", np_id,
      "<br>median: ", median
    ))
    
```

## Methods and source

- Source: Ministry for the Environment, Groundwater quality 1964-2014.
- Files used: `data/mfe-groundwater/groundwater-quality-19642014.gpkg` and `data/mfe-groundwater/groundwater-quality-19642014.txt`.
- License: Creative Commons Attribution 3.0 New Zealand.
- Caveat: indicator fields differ across records and years; this page emphasizes structure and coverage rather than final compliance assessment.

::: {.callout-note collapse="true"}
## Show code

The complete source code is available in folded code chunks above.
:::
