---
title: River Water Quality Nitrogen Trends
toc: true
---

## Headline finding

River nitrogen trend monitoring shows variable trajectories across sites and assessment windows, with trend direction and coverage depending on parameter and period.

## Why this matters for councils

River nitrogen trend indicators help target catchments where deteriorating water quality may require tighter source controls, restoration investment, or more focused monitoring.

## Interactive overview

```{r}
#| warning: false
#| message: false
#| code-fold: true
library(sf)
library(dplyr)
library(plotly)
library(leaflet)

gpkg_path <- "../data/river-water-quality/river-water-quality-nitrogen-trends-1991-2020.gpkg"

layers <- st_layers(gpkg_path)
rwq <- st_read(gpkg_path, layer = layers$name[1], quiet = TRUE)
rwq_tbl <- st_drop_geometry(rwq)

pick_col <- function(df, candidates) {
  hits <- intersect(candidates, names(df))
  if (length(hits) == 0) NA_character_ else hits[1]
}

parameter_col <- pick_col(
  rwq_tbl,
  c("parameter", "Parameter", "indicator", "Indicator", "np_id_name", "np_id", "attribute")
)
period_col <- pick_col(
  rwq_tbl,
  c("period", "Period", "trend_period", "window", "time_window", "assessment_period")
)
trend_col <- pick_col(
  rwq_tbl,
  c("trend", "Trend", "trend_direction", "direction", "Direction", "classification", "trend_class")
)

if (!is.na(trend_col) && !is.na(period_col)) {
  trend_profile <- rwq_tbl |>
    mutate(
      trend_band = as.character(.data[[trend_col]]),
      period_band = as.character(.data[[period_col]])
    ) |>
    filter(!is.na(trend_band), !is.na(period_band), trend_band != "", period_band != "") |>
    count(period_band, trend_band, name = "records")

  plot_ly(
    trend_profile,
    x = ~period_band,
    y = ~records,
    color = ~trend_band,
    type = "bar"
  ) |>
    layout(
      barmode = "stack",
      xaxis = list(title = "Assessment period"),
      yaxis = list(title = "Number of site records"),
      legend = list(title = list(text = "Trend direction"))
    )
} else if (!is.na(trend_col)) {
  trend_profile <- rwq_tbl |>
    mutate(trend_band = as.character(.data[[trend_col]])) |>
    filter(!is.na(trend_band), trend_band != "") |>
    count(trend_band, name = "records")

  plot_ly(
    trend_profile,
    x = ~reorder(trend_band, records),
    y = ~records,
    type = "bar",
    marker = list(color = "#2b7a78")
  ) |>
    layout(
      xaxis = list(title = "Trend direction"),
      yaxis = list(title = "Number of site records"),
      margin = list(b = 120)
    )
} else {
  fallback_col <- ifelse(is.na(parameter_col), names(rwq_tbl)[1], parameter_col)

  coverage_profile <- rwq_tbl |>
    mutate(group_name = as.character(.data[[fallback_col]])) |>
    filter(!is.na(group_name), group_name != "") |>
    count(group_name, sort = TRUE, name = "records") |>
    slice_head(n = 12)

  plot_ly(
    coverage_profile,
    x = ~reorder(group_name, records),
    y = ~records,
    type = "bar",
    marker = list(color = "#2b7a78")
  ) |>
    layout(
      xaxis = list(title = "Indicator field"),
      yaxis = list(title = "Number of records"),
      margin = list(b = 160)
    )
}
```

```{r}
#| warning: false
#| message: false
#| code-fold: true

rwq_map <- rwq


# convert dataframe to sf object: coerce coords to numeric and remove invalid rows
rwq_map$long_num <- as.numeric(rwq_map$long)
rwq_map$lat_num <- as.numeric(rwq_map$lat)
# drop rows with NA coords (including non-numeric coercion results)
rwq_map <- rwq_map[!is.na(rwq_map$long_num) & !is.na(rwq_map$lat_num), ]
rwq_map <- st_as_sf(rwq_map, coords = c("long_num", "lat_num"), crs = 4326, remove = TRUE)


if (nrow(rwq_map) > 0) {
  rwq_map <- suppressWarnings(st_transform(rwq_map, 4326))
}

rwq_map_tbl <- st_drop_geometry(rwq_map)

site_col <- pick_col(rwq_map_tbl, c("site_name", "site", "s_id", "site_id", "location", "name"))
parameter_col <- pick_col(rwq_map_tbl, c("parameter", "Parameter", "indicator", "np_id_name", "np_id"))
period_col <- pick_col(rwq_map_tbl, c("period", "Period", "trend_period", "window", "assessment_period"))
trend_col <- pick_col(rwq_map_tbl, c("trend", "Trend", "trend_direction", "direction", "classification", "trend_class"))
value_col <- pick_col(rwq_map_tbl, c("median", "result", "Result", "value", "Value", "slope", "Sen_slope", "sen_slope"))

popup_fields <- unique(na.omit(c(site_col, parameter_col, period_col, trend_col, value_col)))

if (length(popup_fields) == 0) {
  popup_text <- rep("River monitoring site", nrow(rwq_map_tbl))
} else {
  popup_text <- apply(
    rwq_map_tbl[, popup_fields, drop = FALSE],
    1,
    function(x) {
      vals <- ifelse(is.na(x) | x == "", "NA", as.character(x))
      paste0(names(x), ": ", vals, collapse = "<br>")
    }
  )
}

leaflet(rwq_map) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addCircleMarkers(
    radius = 4,
    stroke = FALSE,
    fillOpacity = 0.55,
    color = "#ff7a59",
    popup = ~paste0(
      "measure: ", "measure",
      "<br>units: ", units,
      "<br>median: ", median,
      "<br>sen_lci: ", "sen_lci",
      "<br>trend: ", "trend_confidence"
    )
  )
```

## Methods and source

- Source: Ministry for the Environment, River water quality nitrogen trends (1991-2020).
- Files used: `data/river-water-quality/river-water-quality-nitrogen-trends-1991-2020.gpkg` and `data/river-water-quality/river-water-quality-nitrogen-trends-1991-2020.txt`.
- License: Creative Commons Attribution 4.0 International.
- Caveat: trend and parameter fields can vary by schema updates; this page emphasizes exploratory structure and pattern coverage rather than formal compliance assessment.

::: {.callout-note collapse="true"}
## Show code

The complete source code is available in folded code chunks above.
:::